### 2024-06-17 sections/4-results.tex
shrunk to 700 lines

Since the time complexity of CAKES algorithms scales with the LFD of the dataset, we examine the LFD of each dataset we used for benchmarks.

Figure~\ref{fig:results:lfd-plots} illustrates the trends in LFD for (the non-augmented versions of) Fashion-Mnist, Glove-25, Sift, Random, Silva 18S, and RadioML.
The horizontal axis denotes the depth in the cluster tree, and the vertical axis denotes the LFD (as calculated by Equation~\ref{eq:methods:lfd-simplified}) of clusters at that depth.
We plot lines for the 5$^{th}$, 25$^{th}$, 50th, 75$^{th}$ and 95$^{th}$ percentiles of LFD, as well as the minimum and maximum LFD at each depth.
In order to have the plots best reflect the distribution of LFDs across the entire \textit{dataset}, in computing percentiles, we count each cluster as many times as its cardinality.
In other words, if, for some dataset, the 95$^{th}$ percentile of LFD at depth 40 is 3, this means that 95\% of the points in clusters at depth 40 belong to a cluster whose LFD is at most 3.

Figure~\ref{fig:results:fashion-mnist-lfd} shows the LFD by depth for Fashion-Mnist.
We observe that until about depth 5, the LFD is low, as the 95$^{th}$ percentile (orange plotted line) is less than 4, and the median (red line) is just above 2.
For depths 15 through 25, we observe that the LFD increases, with the 95$^{th}$ percentile (orange plotted line) slightly less than 6, and the median near 3.
Finally, for depths 25 through the maximum depth, we observe that the LFD decreases again, as the 95$^{th}$ percentile is between 3 and 4, and the median is less than 2.

@@

Figures~\ref{fig:results:fashion-mnist-scaling},~\ref{fig:results:glove-25-scaling},~and~\ref{fig:results:sift-scaling} show the scaling behavior of CAKES algorithms and existing algorithms on augmented versions of Fashion-Mnist under Euclidean distance,
Glove-25 under cosine distance, and
Sift under Euclidean distance.

Figure~\ref{fig:results:random-scaling} shows the scaling behavior of CAKES algorithms and existing algorithms on a completely randomly generated dataset with the same cardinality and dimensionality as Sift (1 million points in 128 dimensions).
Figures~\ref{fig:results:silva-scaling} and ~\ref{fig:results:radioml-scaling} show the scaling behavior of CAKES on Silva and RadioML respectively.
For these datasets, we took random sub-samples of the full datasets instead of augmenting them to higher cardinalities.

The horizontal axis in each figure shows the cardinality of the dataset augmented with synthetic points (see Section \ref{sec:methods:synthetic-data}).
The left-most point on each line is at the cardinality of the original dataset without any synthetic augmentation.
The vertical axis denotes throughput in queries per second (higher is better).
Both axes are on a logarithmic scale.
In this section, we report results only for $k$-NN search with $k = 10$, but similar plots for $k = 100$ can be found in the Supplement.
For HNSW and ANNOY, we report the recall for each measurement in the plots. For algorithms which exhibit recall greater than $0.9995$, we do not report the recall in the plots, but we do report it in the tables below.

Table~\ref{tab:results:qps-and-recall} show the throughput and recall of CAKES's algorithms at each augmented cardinality for Fashion-Mnist, Glove-25, Sift, and Random. 
Though the plots in Figure~\ref{fig:results:scaling-plots} present results for each of CAKES's three algorithms separately, the results in the CAKES column in these tables represent the fastest CAKES algorithm at that dataset and cardinality only.
We used our auto-tuning approach (see Section~\ref{sec:methods:auto-tuning}) for each new tree (i.e. at each cardinality), and this approach was always able to select the fastest algorithm for each dataset at each cardinality.
Since we also allow for tuning hyper-parameters for the other algorithms, and we allow for different sets of hyper-parameters at each cardinality, it is a fair comparison for these tables to only list the performance of the tuned CAKES algorithm.
When reporting recall, we use $1.000*$ to denote that the recall is imperfect, but rounds to $1.000$ when we consider only three decimal places.

Figure~\ref{fig:results:scaling-plots} shows that for Fashion-Mnist, Glove-25, and Sift, as cardinality increases, the CAKES algorithms (Depth-First Sieve in blue, Repeated $\rho$-NN in green, and Breadth-First Sieve in purple) become faster than our Rust implementation of na\"{i}ve linear search (in orange).
Though we observe this trend, we note that the exact cardinality at which CAKES's algorithms overtake linear search differs by dataset. For Fashion-Mnist, CAKES begins outperforming linear search starting at a cardinality near $10^5$, while for Glove-25 and Sift, this happens near $10^6$ and $10^7$ respectively. 
Which one of the three CAKES algorithms is fastest also differs by dataset.
For Fashion-Mnist, Depth-First Sieve is consistently fastest, while for Glove-25, the fastest is Repeated $\rho$-NN, and for Sift, Breadth-First Sieve. 
Additionally, on all three datasets, Depth-First Sieve and Breadth-First Sieve appear to have performance which is \textit{constant} in the cardinality of the dataset. 
With Glove-25, Repeated $\rho$-NN exhibits similar constant scaling. 
We also observe that on these three datasets, for nearly all cardinalities, all three of the CAKES algorithms are faster than FAISS-Flat (in brown), and that at some cardinality, CAKES's algorithms become faster than FAISS-IVF (in pink). 
For Fashion-Mnist, CAKES becomes faster than FAISS-IVF near cardinality $10^5$, whereas for Glove-25 and Sift, this happens near cardinality $10^7$. 
On all three of these datasets, HNSW (in gray) and ANNOY (in yellow) are faster than CAKES's algorithms for all cardinalities; however, CAKES exhibits perfect or near-perfect recall on each dataset, while HNSW and ANNOY exhibit much lower recall, as shown in Table~\ref{tab:datasets:summary}.
While recall for CAKES's does \emph{not} degrade with cardinality, recall for HNSW and ANNOY does degrade with cardinality. 
At a cardinality multiplier as low as eight, HNSW and ANNOY have recall of $0.525$ and $0.857$ respectively for Fashion-Mnist, 0.607 and 0.832 for Glove-25, and 0.782 and 0.686 on Sift. 
In contrast, CAKES has perfect recall on Fashion-Mnist and Sift (because the distance function used with these datasets is a metric), and near-perfect recall on Glove-25 (because the cosine distance function is not a metric).

In contrast with the results on the ANN Benchmark datasets reported above, with the Random dataset, as seen in Figure~\ref{fig:results:random-scaling}, we observe that the three CAKES algorithms are the slowest out of all algorithms, at all cardinalities.
As with the real datasets, HNSW and ANNOY are the fastest algorithms, and CAKES exhibits perfect recall at all cardinalities.
HNSW and ANNOY exhibit \textit{much} lower recall on this random dataset than on any of the ANN benchmark datasets;
in particular, with a multiplier of 1, HNSW and ANNOY have recall as low as 0.060 and 0.028 respectively, as reported in Table~\ref{tab:results:qps-and-recall}.

For Silva and RadioML, we benchmarked only CAKES's algorithms because HNSW, ANNOY and FAISS support neither the required distance functions nor, in the case of RadioML, complex-valued data.
Due to the massive sizes of these datasets and challenges in generating plausible augmentations, we took random sub-samples ranging up to the entirety of the dataset--rather than augmented versions of the dataset--to examine how performance scales with cardinality.
With Silva, as shown in Figure~\ref{fig:results:silva-scaling}, we observe that for all algorithms, throughput initially seems to linearly decrease as cardinality increases, but that it seems to begin levelling off near cardinality $10^5$.
Until cardinality near $10^4$, Depth-First Sieve is the fastest CAKES algorithm, but for larger cardinalities, Repeated $\rho$-NN is the fastest CAKES algorithm.
For RadioML, as shown in Figure~\ref{fig:results:radioml-scaling}, we observe that throughput declines nearly linearly, and that the three CAKES algorithms exhibit virtually indistinguishable performance for all cardinalities.
In particular, throughput of CAKES's algorithms is identical within three significant figures for all cardinalities.

>>>>

Since the time complexity of CAKES algorithms scales with the LFD of the dataset, we examine the LFD of each dataset we used for benchmarks.
Figure~\ref{fig:results:lfd-plots} illustrates the trends in LFD for Fashion-Mnist, Glove-25, Sift, Random, Silva 18S, and RadioML.
The horizontal axis denotes depth in the cluster tree, and the vertical axis denotes the LFD of clusters at that depth.
We plot lines for the 5$^{th}$, 25$^{th}$, 50th, 75$^{th}$ and 95$^{th}$ percentiles of LFD, as well as the minimum and maximum LFD at each depth.
In order to have the plots best reflect the distribution of LFDs across the entire \textit{dataset}, we count each cluster as many times as its cardinality.
In other words, if, for some dataset, the 95$^{th}$ percentile of LFD at depth 40 is 3, this means that 95\% of the points in clusters at depth 40 belong to a cluster whose LFD is at most 3.
Figure~\ref{fig:results:fashion-mnist-lfd} shows the LFD by depth for Fashion-Mnist.
We observe that until about depth 5, LFD is low, as the 95$^{th}$ percentile (orange plotted line) is less than 4, and the median (red line) is just above 2.
For depths 15 through 25, we observe that LFD increases, with the 95$^{th}$ percentile (orange plotted line) slightly less than 6, and the median near 3.
Finally, for depths 25 through the maximum depth, we observe that the LFD decreases again, as the 95$^{th}$ percentile is between 3 and 4, and the median is less than 2.

@@

Figures~\ref{fig:results:fashion-mnist-scaling},~\ref{fig:results:glove-25-scaling},~and~\ref{fig:results:sift-scaling} show the scaling behavior of CAKES algorithms and existing algorithms on augmented versions of Fashion-Mnist under Euclidean distance,
Glove-25 under cosine distance, and
Sift under Euclidean distance.
Figure~\ref{fig:results:random-scaling} shows the scaling behavior of CAKES algorithms and existing algorithms on a completely randomly generated dataset with the same cardinality and dimensionality as Sift (1 million points in 128 dimensions).
Figures~\ref{fig:results:silva-scaling} and ~\ref{fig:results:radioml-scaling} show the scaling behavior of CAKES on Silva and RadioML respectively.
For these datasets, we took random sub-samples of the full datasets instead of augmenting them to higher cardinalities.
The horizontal axis in each figure shows the cardinality of the dataset augmented with synthetic points (see Section \ref{sec:methods:synthetic-data}).
The left-most point on each line is at the cardinality of the original dataset.
The vertical axis denotes throughput in queries per second.
Both axes are on a logarithmic scale.
In this section, we report results only for $k$-NN search with $k = 10$, but similar plots for $k = 100$ can be found in the Supplement.
For HNSW and ANNOY, we report the recall for each measurement in the plots. For algorithms which exhibit recall greater than $0.9995$, we do not report the recall in the plots, but we do report it in the tables below.

Table~\ref{tab:results:qps-and-recall} show the throughput and recall of CAKES's algorithms at each augmented cardinality for Fashion-Mnist, Glove-25, Sift, and Random.
Though the plots in Figure~\ref{fig:results:scaling-plots} present results for each of CAKES's three algorithms separately, the results in the CAKES column in these tables represent the fastest CAKES algorithm at that dataset and cardinality only.
We used our auto-tuning approach (see Section~\ref{sec:methods:auto-tuning}) for each new tree (i.e. at each cardinality), and this approach was always able to select the fastest algorithm for each dataset at each cardinality.
Since we also allow for tuning hyper-parameters for the other algorithms, and we allow for different sets of hyper-parameters at each cardinality, it is a fair comparison for these tables to only list the performance of the tuned CAKES algorithm.
When reporting recall, we use $1.000*$ to denote that the recall is imperfect, but rounds to $1.000$.

Figure~\ref{fig:results:scaling-plots} shows that for Fashion-Mnist, Glove-25, and Sift, as cardinality increases, the CAKES algorithms (Depth-First Sieve in blue, Repeated $\rho$-NN in green, and Breadth-First Sieve in purple) become faster than our Rust implementation of na\"{i}ve linear search (in orange).
Though we observe this trend, we note that the exact cardinality at which CAKES's algorithms overtake linear search differs by dataset. For Fashion-Mnist, CAKES begins outperforming linear search starting at a cardinality near $10^5$, while for Glove-25 and Sift, this happens near $10^6$ and $10^7$ respectively.
Which one of the three CAKES algorithms is fastest also differs by dataset.
For Fashion-Mnist, Depth-First Sieve is consistently fastest, while for Glove-25, the fastest is Repeated $\rho$-NN, and for Sift, Breadth-First Sieve.
On all three datasets, Depth-First Sieve and Breadth-First Sieve appear to have performance which is \textit{constant} in the cardinality of the dataset.
With Glove-25, Repeated $\rho$-NN exhibits similar constant scaling.
We also observe that on these three datasets, for nearly all cardinalities, all three of the CAKES algorithms are faster than FAISS-Flat (in brown), and that at some cardinality, CAKES's algorithms become faster than FAISS-IVF (in pink).
For Fashion-Mnist, CAKES becomes faster than FAISS-IVF near cardinality $10^5$, whereas for Glove-25 and Sift, this happens near cardinality $10^7$.
On all three of these datasets, HNSW (in gray) and ANNOY (in yellow) are faster than CAKES's algorithms for all cardinalities; however, CAKES exhibits perfect or near-perfect recall on each dataset, while HNSW and ANNOY exhibit much lower recall, as shown in Table~\ref{tab:datasets:summary}.
While recall for CAKES's does \emph{not} degrade with cardinality, recall for HNSW and ANNOY degrades with cardinality.
At a cardinality multiplier as low as eight, HNSW and ANNOY have recall of $0.525$ and $0.857$ respectively for Fashion-Mnist, 0.607 and 0.832 for Glove-25, and 0.782 and 0.686 on Sift.
In contrast, CAKES has perfect recall on Fashion-Mnist and Sift (where the distance function is a metric), and near-perfect recall on Glove-25 (cosine distance is not a metric).

In contrast with the results on the ANN Benchmark datasets reported above, with the Random dataset, as seen in Figure~\ref{fig:results:random-scaling}, we observe performs quite slowly.
As with the real datasets, HNSW and ANNOY are the fastest algorithms, and CAKES exhibits perfect recall at all cardinalities.
HNSW and ANNOY exhibit \textit{much} lower recall on this random dataset than on any of the ANN benchmark datasets;
in particular, with a multiplier of 1, HNSW and ANNOY have recall as low as 0.060 and 0.028 respectively, as reported in Table~\ref{tab:results:qps-and-recall}.

For Silva and RadioML, we benchmarked only CAKES's algorithms because HNSW, ANNOY and FAISS support neither the required distance functions nor, in the case of RadioML, complex-valued data.
Due to the massive sizes of these datasets and challenges in generating plausible augmentations, we took random sub-samples ranging up to the entirety of the dataset--rather than augmented versions of the dataset--to examine how performance scales with cardinality.
With Silva, as shown in Figure~\ref{fig:results:silva-scaling}, we observe that for all algorithms, throughput initially seems to linearly decrease as cardinality increases, but that it seems to begin levelling off near cardinality $10^5$.
Until cardinality near $10^4$, Depth-First Sieve is the fastest CAKES algorithm, but for larger cardinalities, Repeated $\rho$-NN is the fastest CAKES algorithm.
For RadioML, as shown in Figure~\ref{fig:results:radioml-scaling}, we observe that throughput declines nearly linearly, and that the three CAKES algorithms exhibit virtually indistinguishable performance.
In particular, throughput of CAKES's algorithms is identical within three significant figures.

### 2024-06-17 sections/5-discussion.tex
shrunk to 700 lines

We have presented CAKES, a set of three algorithms for fast $k$-NN search that are generic over a variety of distance functions.
CAKES's algorithms are exact when the distance function in use is a metric (see definition in~\ref{sec:methods}).
Even under cosine distance, which is not a metric, CAKES's algorithms exhibit nearly perfect recall (i.e. always greater than $0.9995$).

CAKES's algorithms are designed to be most effective when the data uphold the manifold hypothesis, or in other words, when the data are constrained to a low-dimensional manifold even when embedded in a high-dimensional space.
As a consequence, these algorithms do not perform well on data with random distributions, because of the absence of a manifold structure.
In Figure~\ref{fig:results:lfd-plots}, we show the extent to which each dataset we tested exhibits a manifold structure, as quantified by the percentiles of LFD.
As expected, our results from Figure~\ref{fig:results:scaling-plots} indicate that CAKES's algorithms scale sub-linearly with cardinality on datasets with low LFD, and scale linearly on datasets with high LFD.

As seen in Figures~\ref{fig:results:random-lfd}~and~\ref{fig:results:sift-lfd}, the LFD of the Random dataset is much higher than that of Sift, even though both datasets have the same cardinality and embedding dimension. 
This is unsurprising, given that the Random dataset is, by definition, uniformly distributed, while Sift is not.
On the Random dataset, CAKES's algorithms' speed decreases linearly as the multiplier increases, while with Sift, Depth-First Sieve and Breadth-First Sieve both exhibit nearly constant throughput.
While Repeated $\rho$-NN does not exhibit near-constant throughput with Sift, the throughput decreases much more slowly than it does with the Random dataset.
Since these two datasets have the same cardinality and embedding dimension, the aforementioned discrepancies highlight how manifold structure affects algorithm performance.
We emphasize that even though CAKES's algorithms are slower on the Random dataset, their recalls remain perfect at all multipliers.

Figure~\ref{fig:results:scaling-plots} also shows that HNSW and ANNOY have near constant throughput as the multiplier increases, however, their recall continues to decrease as the multiplier increases.
This, coupled with the prohibitively increasing cost (in terms of memory and time) of building the indices for HNSW and ANNOY, makes these algorithms unsuitable for real-world applications in which the dataset cardinalities are growing exponentially.
CAKES's tree is orders of magnitude faster and less memory-intensive to build and CAKES's algorithms, while a little slower than HNSW and ANNOY, still show near constant scaling with cardinality in addition to having perfect recall.

@@

As the previous paragraph suggests, the fastest CAKES algorithm varies by dataset.
For example, on Glove-25, Repeated $\rho$-NN exhibits the highest throughput of any CAKES algorithm, despite having the lowest throughput of the three on Sift and Fashion-Mnist.
On Silva, Repeated $\rho$-NN is comparable to the other CAKES algorithms at low multipliers, but becomes the fastest at the higher multipliers.
When we view these results in light of the LFD plots in Figure~\ref{fig:results:lfd-plots}, we realize that Repeated $\rho$-NN appears to be the fastest CAKES algorithm on datasets where a large proportion of the data have a very low LFD, such as Glove-25 and Silva.
This trend is unsurprising, given that Repeated $\rho$-NN relies on a low LFD around the query in order to quickly estimate the correct radius for $k$ hits.
These findings suggest that Repeated $\rho$-NN is the most ``sensitive'' to the manifold structure of the dataset.
Exploring this sensitivity is a potential avenue for future investigation, as it remains to be determined at what LFD Repeated $\rho$-NN transitions away from being the fastest CAKES algorithm, and whether this transition is sharp.

While the fastest CAKES algorithm varies by dataset, we observe that on all three ANN-benchmark datasets (Fashion-Mnist, Glove-25, and Sift), Depth-First Sieve and Breadth-First Sieve show nearly constant throughput as the cardinality increases.
This observation warrants further investigation, but it is especially promising given that Depth-First Sieve and Breadth-First Sieve consistently outperform linear search for high cardinalities.
Finally, we emphasize that the variation in performance of our algorithms across different datasets and cardinalities supports our use of an auto-tuning function (as discussed in Section~\ref{sec:methods:auto-tuning}) to select the best algorithm for a given dataset.
Future work is needed to better characterize those datasets for which each algorithm performs best, but our results suggest that a more sophisticated auto-tuning function could be developed to select the fastest algorithm based on the properties of the dataset.

>>>>

We have presented CAKES, a set of three algorithms for fast $k$-NN search that are generic over a variety of distance functions.
CAKES's algorithms are exact when the distance function in use is a metric (see definition in~\ref{sec:methods}).
Even under cosine distance, which is not a metric, CAKES's algorithms exhibit nearly perfect recall.
CAKES's algorithms are designed to be most effective when the data uphold the manifold hypothesis, or in other words, when the data are constrained to a low-dimensional manifold even when embedded in a high-dimensional space.
As a consequence, these algorithms do not perform well on data with random distributions, because of the absence of a manifold structure.
In Figure~\ref{fig:results:lfd-plots}, we show the extent to which each dataset we tested exhibits a manifold structure, as quantified by the percentiles of LFD.
As expected, our results from Figure~\ref{fig:results:scaling-plots} indicate that CAKES's algorithms scale sub-linearly with cardinality on datasets with low LFD, and scale linearly on datasets with high LFD.

As seen in Figures~\ref{fig:results:random-lfd}~and~\ref{fig:results:sift-lfd}, the LFD of the Random dataset is much higher than that of Sift, even though both datasets have the same cardinality and dimensionality.
This is unsurprising, given that the Random dataset is uniformly distributed, while Sift is not.
On the Random dataset, CAKES's speed decreases linearly as the multiplier increases, while with Sift, Depth-First Sieve and Breadth-First Sieve both exhibit nearly constant throughput.
While Repeated $\rho$-NN does not exhibit near-constant throughput with Sift, throughput decreases much more slowly than it does with the Random dataset.
Since these two datasets have the same cardinality and embedding dimension, the aforementioned discrepancies highlight how manifold structure affects algorithm performance.
We emphasize that although CAKES's algorithms are slower on the Random dataset, their recalls remain perfect.
Figure~\ref{fig:results:scaling-plots} also shows that HNSW and ANNOY have near-constant throughput as the multiplier increases, but their recall continues to decrease as the multiplier increases.
This, coupled with the increasing time and space cost of building the indices for HNSW and ANNOY, makes these algorithms unsuitable for real-world applications in which the datasets are growing exponentially.
CAKES's tree is orders of magnitude faster and less memory-intensive to build, and CAKES's algorithms, while slower than HNSW and ANNOY, still show near-constant scaling as well as perfect recall.

@@

As the previous paragraph suggests, the fastest CAKES algorithm varies by dataset.
For example, on Glove-25, Repeated $\rho$-NN exhibits the highest throughput of any CAKES algorithm, despite having the lowest throughput of the three on Sift and Fashion-Mnist.
On Silva, Repeated $\rho$-NN is comparable to the other CAKES algorithms at low multipliers, but becomes the fastest at the higher multipliers.
When we view these results in light of the LFD plots in Figure~\ref{fig:results:lfd-plots}, we realize that Repeated $\rho$-NN appears to be the fastest CAKES algorithm on datasets where a large proportion of the data have a very low LFD, such as Glove-25 and Silva.
This trend is unsurprising, given that Repeated $\rho$-NN relies on a low LFD around the query in order to quickly estimate the correct radius for $k$ hits.
These findings suggest that Repeated $\rho$-NN is the most ``sensitive'' to the manifold structure of the dataset.
Exploring this sensitivity is a potential avenue for future study, as it remains to be seen at what LFD Repeated $\rho$-NN transitions away from being the fastest CAKES algorithm, and whether this transition is sharp.

While the fastest CAKES algorithm varies by dataset, we note that on all three ANN-benchmark datasets (Fashion-Mnist, Glove-25, and Sift), Depth-First Sieve and Breadth-First Sieve show nearly constant throughput as the cardinality increases.
This observation warrants further investigation, but it is especially promising given that both algorithms consistently outperform linear search for high cardinalities.
Finally, we emphasize that the variation in performance of our algorithms across different datasets and cardinalities supports our use of an auto-tuning function (as discussed in Section~\ref{sec:methods:auto-tuning}) to select the best algorithm for a given dataset.
Future work is needed to better characterize those datasets for which each algorithm performs best, but our results suggest that a more sophisticated auto-tuning function could be developed to select the fastest algorithm based on the properties of the dataset.
